\addtocounter{chapter}{-1}
\chapter{Homework assignments}


It is better to think about all the problems, but you do not have to solve \emph{all} of them.
If a problem is solved, you do not have to write its solutions, but try sketch it.

\section{Due Tue Jan 21}

Exercises: \sout{\ref{ex:almost-min},} \ref{ex:non-contracting-map}, \ref{ex:no-geod}, \sout{\ref{ex:compact=>complete},} \ref{exercise from BH}, \ref{ex:Hausdorff-bry}.

\section{Due Tue Jan 28}

Exercises: \ref{ex:almost-min},  \ref{ex:compact=>complete}, \ref{ex:Huas-perimeter-area}, \ref{ex:GH-po}, \ref{pr:doubling}, \ref{pr:under:if}.

\section{Due Tue Feb 4}
Exercises: 
\ref{ex:compact-length}, 
\ref{pr:under:only-if}, 
\sout{\ref{ex:GH-SC},}
\sout{\ref{ex:sphere-to-ball},}
\ref{ex:ultrapower}, 
\ref{ex:two-geodesics-in-ultrapower}.

\section{Due Tue Feb 11}

Finish exercises \ref{ex:compact-length} , \ref{pr:under:only-if}, \ref{ex:GH-SC}, \ref{ex:sphere-to-ball}.

\noindent
Exercises: \ref{ex:lim(tree)}, \ref{ex:Asym(Lob)}, \ref{ex:geodesics-urysohn}, \ref{ex:sphere-in-urysohn}.

\section{Due Tue Feb 18}

Exercises: \ref{ex:compact-extension}, \ref{ex:+-c}, \ref{ex:ultrametric}, \ref{ex:injective-spaces}, \ref{ex:tripod+square}, \ref{ex:4-on-a-line}.

\noindent Write down a solution of at least one of the exercises.

\section{Due Tue Feb 25}

Finish Exercise \ref{ex:tripod+square:square}.
Prepare questions for review on Tuesday.

\section{Due Tue Mar 3}

Exercises: \ref{ex:sba-2+2-short}, \ref{ex:(3+1)-expanding}, \ref{ex:CAT+CBB}, \ref{ex:product-CBB}, \sout{\ref{ex:CBB-geodesic},} \ref{ex:fat-triangle}.

\noindent Write down a solution of at least one of the exercises.

\section{Due Tue Mar 17}

Exercises: \ref{ex:tringle-inq-angles},
\ref{ex:CBB-geodesic},
\ref{ex:convex-dist},
\ref{ex:reshetnyak-doubling},
\ref{ex:supporting-planes},
\ref{ex:centrally-simmetric-walls}.

\noindent Write down a solution of at least one of the exercises.

\section{Due Tue Mar 24}

Exercises: 
\ref{ex:contractible},
\ref{ex:convex-nbhd},
\ref{ex:closest-point},
\ref{cor:balls:dim=1},
\ref{ex:null-homotopic},
\sout{\ref{ex:branching-cover}.}

 Write down as many solutions as you can; email it to Zetian Yan (zxy5156) + cc to me (aqp6).

\section{Due Tue Mar 31}

Exercises: 
\ref{ex:branching-cover},
\ref{ex:tan(CAT)isCAT},
\ref{ex:tan(CAT)is-length},
%\ref{ex:product-cone},
\ref{ex:unique-geod=CAT},
\ref{ex:flag>=pi/2},
\ref{ex:tree}.

Write down as many solutions as you can; email it to Zetian Yan (zxy5156) + cc to me (aqp6).

\section{Due Tue Apr 7}

Exercises: 
\ref{ex:CAT-mnfld=>ext.geod},
\ref{ex:locally-convex},
\ref{ex:geod-circle},
\ref{ex:flag-aspherical},
\ref{ex:example-pi_infty-new},
\ref{ex:cube-infty=>cube-2}.

Write down as many solutions as you can; email it to Zetian Yan (zxy5156) + cc to me (aqp6).

\section{Due Tue Apr 14}

Exercises: 
\ref{ex:geod-CBA},
%\ref{prop:two-hull-open},
%\ref{ex:chopping-triangle},
\ref{ex:concave-triangle},
\ref{ex:two-planes},
\ref{ex:hemisphere},
\ref{ex:inner-support},
\ref{ex:convex+saddle+broken=>PL}.

Write down as many solutions as you can; email it to Zetian Yan (zxy5156) + cc to me (aqp6).

\section*{Remark}
Each working day I will check email before 15:00 and will appear online if you ask (it is easy for me --- do not hesitate to ask).
We will meet regular hours online (as we did before).
