

\parbf{\ref{ex:gross}.}
If such a number does not exist, then the ranges of average distance functions have empty intersection.
Since $\spc{X}$ is a compact length-metric space, the range of any continuous function on $\spc{X}$ is a closed interval.
By 1-dimensional Helly's theorem, there is a pair of such range intervals that do not intersect.
That is, for two point-arrays $(x_1,\dots,x_n)$ and $(y_1,\dots,y_m)$
and their average distance functions 
\[f(z)=\tfrac1n\cdot\sum_i|x_i-z|_{\spc{X}}\quad\text{and}\quad h(z)=\tfrac1m\cdot\sum_j|y_j-z|_{\spc{X}},\] we have 
$$\min\set{f(z)}{z\in \spc{X}}>\max\set{h(z)}{z\in \spc{X}}.\leqno({*})$$

Note that 
$$\tfrac1m\cdot\sum_j f(y_j)=\tfrac1{m\cdot n}\cdot\sum_{i,j}|x_i-y_j|_{\spc{X}}=\tfrac1n\cdot\sum_i h(x_i);$$
that is, the average value of $f(y_j)$ coincides with the average value of $h(x_i)$, 
which contradicts $({*})$.

\parit{Remarks.}
In fact the value $\ell$ is uniquely defined;
it is called the \index{rendezvous value}\emph{rendezvous value} of ${\spc{X}}$.
This is a result of Oliver Gross \cite{gross}.

\parbf{\ref{ex:wasserstein}.}
Choose a finite $\eps$-net $F\subset\spc{X}$.
Show that the space $P_F$ of probability measures with support in $F$ is a compact net in $\Wass_1\spc{X}$.
Observe that $\Wass_1\spc{X}$ is complete; 
by \ref{ex:compact-net}, it follows that $\Wass_1\spc{X}$ is compact.

Show that 

Choose an integer $n$.
Consider the set of probability measures $P_n$ of the form 
\[\tfrac1n\cdot\sum_{i=1}^n\delta_{x_i},\]
where $\delta_{x_i}$ denotes the delta-measure supported at $x_i\in\spc{X}$. 

Show that $P_n$ is a compact subset of $\Wass_1\spc{X}$.
Moreover, for any $\eps>0$ there is $n$ such that $P_n$ is 




%%%%%%%%%%%%%%%%%%%%%%%%%

\parbf{\ref{ex:compact-length}.} By Fréchet lemma (\ref{lem:frechet}) we can identify $\spc{K}$ with a compact subset of $\ell^\infty$.

Denote by $\spc{L}=\Conv\spc{K}$ --- it is defined as the minimal convex closed set in $\ell^\infty$ that contains $\spc{K}$.
(In other words, $\spc{L}$ is the intersection of all convex closed sets that contain $\spc{K}$.)

Observe that $\spc{L}$ is a length space.
It remains to show that $\spc{L}$ is compact.

By construction $\spc{L}$ is a closed subset of $\ell^\infty$, in particular it is a complete space.
By \ref{totally-bounded}, it remains to show that $\spc{L}$ is totally bounded.

Recall that Minkowski sum $A + B$ of two sets $A$ and $B$ in a vector space is defined by
\[A + B = \set{a+b}{a\in A,\ b\in B}.\]
Observe that the Minkowski sum of two convex sets is convex.

Denote by $\bar B_\eps$ the closed $\eps$-ball in $\ell^\infty$ centered at the origin.
Choose a finite $\eps$-net $N$ in $\spc{K}$ for some $\eps>0$.
Note that $P=\Conv N$ is a convex polyhedron, in particular $\Conv N$ is compact.

Observe that $N+\bar B_\eps$ is closed $\eps$-neighborhood of $N$.
It follows that $N+\bar B_\eps\supset K$ and therefore $P+\bar B_\eps\supset \spc{L}$.
In particular $P$ is a $2\cdot\eps$-net in $\spc{L}$;
since $P$ is compact and $\eps>0$ is arbitrary, $\spc{L}$ is totally bounded (see \ref{ex:compact-net}).

\parit{Remark.}
Another solution follows since the injective envelope of a compact space is compact; see \ref{ex:inj=complete-geodesic-contractible:geodesic}, \ref{ex:Inj(compact)}, and \ref{prop:InjX-is-injective}.

\parbf{\ref{ex:geodesics-urysohn}.}
Choose a separable space $\spc{X}$ that has an infinite number of geodesics between a pair of points, say a square will $\ell^\infty$-metric in $\RR^2$.
Apply to $\spc{X}$ universality of Urysohn space (\ref{prop:sep-in-urys}).

\parbf{\ref{ex:compact-extension}.} 
First let us prove the following claim:

\begin{itemize}
\item 
Suppose $f\: K\to\RR$ is an extension function defined on a compact subset $K$ of the Urysohn space $\spc{U}$.
Then there is a point $p\in \spc{U}$ such that 
$\dist{p}{x}{}=f(x)$ for any $x\in K$.
\end{itemize}

Without loss of generality we may assume that $f(x)>0$ for any $x\in K$.
Since $K$ is compact, we may fix $\eps>0$ such that $f(x)>\eps$.

Consider the sequence $\eps_n=\tfrac\eps{100\cdot 2^n}$.
Choose a sequence of $\eps_n$-nets $N_n\subset K$.
Applying universality of $\spc{U}$ recursively, we may choose a point $p_n$ such that $\dist{p_n}{x}{}=f(x)$ for any $x\in N_n$ and $\dist{p_n}{p_{n-1}}{}\z=10\cdot\eps_{n-1}$.
Observe that the sequence $(p_n)$ is Cauchy and its limit $p$ meets 
$\dist{p}{x}{}=f(x)$ for any $x\in K$.

Now, choose a sequence of points $(x_n)$ in $\spc{S}$.
Applying the claim, we may extend the map from $K$ to $K\cup\{x_1\}$, and further to $K\cup\{x_1,x_2\}$, and so on.
As a result we extend the distance-preserving map $f$ to the whole sequence $(x_n)$.
It remains to extend it continuously to the whole space~$\spc{S}$.

\parbf{\ref{ex:sc-urysohn}.}
It is sufficient to show that any compact subspace $\spc{K}$ of Urysohn space can be contracted to a point.

Note that any compact space $\spc{K}$ can be extended to a contractible compact space $\spc{K}'$; for example we may embed $\spc{K}$ into $\ell^\infty$ and pass to its convex hull, as it was done in \ref{ex:compact-length}.

By \ref{thm:compact-homogeneous}, there is an isometric embedding of $\spc{K}'$ that agrees with inclusion of $\spc{K}$.
Since $\spc{K}$ is contractible in $\spc{K}'$, it is contractible in $\spc{U}$.

\parit{A better way.}
One can contract the whole Urysohn space using the following construction.

Note that points in the space $\spc{X}_\infty$ constructed in the proof of \ref{prop:univeral-separable} can be multiplied number $t\in [0,1]$ --- simply multiply each function by $t$.
That defines a map 
\[\lambda_t\:\spc{X}_\infty\to \spc{X}_\infty\]
that scales all distances by factor $t$.
The map $\lambda_t$ can be extended to the completion of $\spc{X}_\infty$, which is isometric to $\spc{U}_d$ (or $\spc{U}$).

Observe that 
the map $\lambda_1$ is the identity  and $\lambda_0$ maps whole space to a single point, say $x_0$ --- this is the only point of $\spc{X}_0$.
Further note that the map $(t,p)\mapsto \lambda_t(p)$ is continuous ---  in particular $\spc{U}_d$ and $\spc{U}$ are contractible.

As a bonus, observe that for any point $p\in \spc{U}_d$ the curve $t\mapsto \lambda_t(p)$ is a geodesic path from $p$ to $x_0$.

\parit{Source:} \cite[(d) on page 82]{gromov-2007}.

\parbf{\ref{ex:sphere-in-urysohn}}; \ref{SHORT.ex:sphere-in-urysohn:sphere} and \ref{SHORT.ex:sphere-in-urysohn:midpoint}.
Observe that $L$ and $M$ satisfy the definition of $d$-Urysohn space and apply the uniqueness (\ref{thm:urysohn-unique}).

\parit{\ref{SHORT.ex:sphere-in-urysohn:homogeneous}.} 
Use \ref{SHORT.ex:sphere-in-urysohn:sphere}, maybe twice.

\parbf{\ref{ex:homogeneous}}; \ref{SHORT.ex:homogeneous:euclidean}.
The euclidean plane is homogeneous in every sense.

\parit{\ref{SHORT.ex:homogeneous:hilbert}.} The ilbert space $\ell^2$ is finite set homogeneous, but not compact set homogeneous, nor countable homogeneous.

\parit{\ref{SHORT.ex:homogeneous:ell-infty}.} The space $\ell^\infty$ is 1-point homogeneous, but not 2-point homogeneous.
Try to show that there is no isometry of $\ell^\infty$ such that
\begin{align*}
(0,0,0,\dots)&\mapsto (0,0,0,\dots),
\\
(1,1,1,\dots)&\mapsto (1,0,0,\dots).
\end{align*}

\parit{\ref{SHORT.ex:homogeneous:ell-1}.}
The space $\ell^1$ is 1-point homogeneous, but not 2-point homogeneous.
Try to show that there is no isometry of $\ell^\infty$ such that
\begin{align*}
(0,0,0,\dots)&\mapsto (0,0,0,\dots),
\\
(2,0,0\dots)&\mapsto (1,1,0,\dots).
\end{align*}
